Blueprint for nuclear spin characterization in color center

### 2023-08-21 article.tex
fisher
a
f

\title{Blueprint for nuclear spin characterization in single color center}% Force line breaks with \\
%\thanks{A footnote to the article title}%

@@

%%%%%%%%%%%%%%%%%%%================= %THEORY
\section{Methods}
\label{sec:theory}
%%%%%%%%%%%%%%%Í%%%%=================
In this section, we provide the theoretical formulas for. In a central electron spin with the electron spin $\boldmath{S}$, we only apply pulses resonant with two of the electron spin sublevels, denoted as $m_s=s_0$ and $m_s = s_1$. These two spin sublevels are separated, due to Zeeman effect or zero field splitting, and form an artificial two-level system with energy splitting of $\omega_a$. Now, we add spin $\frac{1}{2}$nuclear spins to the system (for considering P1 centers, one has to add nuclear quadrupole to the presented Hamiltonian.). The Hamiltonian of a central electron spins system surrounded by nuclear spins in the secular approximation in the rotating frame of the applied microwave $\omega_{mw}$ (considering rotating wave approximation) can be written as:
\begin{align}
	H = \Delta S_z + \sum_k (\omega_L^{(k)}  + A_{zz}^{(k)} S_z) I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
	\label{eq:H}
\end{align}
Where $\Delta = \omega_a - \omega_{mw}$ is the detuning, $\omega_L^{(k)} =\gamma_n^{(k)} B$ is nuclear Rabi frequency with $\gamma_n^{(k)}$ nuclear spin gyromagnetic ratio and $B$ external magnetic field, $A_{zz}, A_{zx}$ parallel and perpendicular secular component of hyperfine tensor. This Hamiltonian results that depending on the electron spin sublevel $s_i$ ($i=0,1$), the nuclear spin rotate with the resonance frequency of $\omega_i = \sqrt{(\omega_L + s_i A_{zz})^2+ (s_i A_{zx})^2}$, $i=0,1$ along the axes $\vec{n}_i =(\frac{s_i A_{zx}}{\omega_i}, 0, \frac{\omega_L + s_i A_{zz}} {\omega_i})$. It is assumed that the pulse duration $t_p$ are short enough ($\omega_L t_p \ll 1$) that the nuclear spin dynamics during this time is negligible. The most simplest characterization sequence is Ramsey. \\ The Ramsey signal can be obtained as follows:
	\expval{\sigma_z}_{Ram} = \cos(\Delta \tau) \prod_{k=1}^{n} \Big[ \cos(\frac{\omega_0 \tau}{2}) \cos(\frac{\omega_1 \tau}{2})+ \notag \\
	(\vec{n_0}\cdot \vec{n_1})\sin(\frac{\omega_0 \tau}{2}) \sin(\frac{\omega_1 \tau}{2}) \Big]^{(k)}
	\textcolor{gray}{e^{-\frac{\tau}{T_2^\ast}}}
We do not take any relaxation process into account and the gray exponential decay is put by hand to remember the relaxation time scale. Ramsey signal decay with electron spin $T_2^\ast$, so it can be used to observe the core of the defect. Ramsey sequence is sensitive to detuning since the electron spin is not refocused, however it is possible to sense nuclear spins with vanishing perpendicular hyperfine coupling.\\

@@

		\includegraphics[width=0.8\textwidth]{pict/fig3.png}

@@

We want to make a sequence that is limited by T1 instead of T2. The idea of 5-pulse ESEEM \cite{kasumaj20085} is that to create an entanglement (Hahn-echo) before the free evolution (Ramsey). In other words, change the initial density matrix for the Ramsey sequence from such that entanglement already exist in the electron polarization terms of the density matrix. Fig \ref{fig:seqs} shows the sequence, which can be interpreted as two Hahn Echos that are separated by a long free evolution $T\gg T_2^\ast$ such that the coherence of the electron spin vanishes. The analytical formula to this sequence is provided in the appendix \ref{app:5pESEEM}. During the free evolution time, each nuclear spins oscillates with either of two resonance frequencies obtained from eq. \ref{eq:H}.
\begin{figure*}[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{pict/fig4.png}
		\caption{a. FFT of ESEEM signal b. a zoom into weakly coupled nuclear spins c. Enhancing weakly coupled nuclear spins by having 72 pulses in each entangling period. $\tau$ is set to the Larmor bright spot.}
		\label{fig:4}
	\end{center}
\end{figure*}
Fig \ref{fig:4} shows the furrier transform of the signal. Zooming into area close to the Larmor frequency reveals the weakly coupled nuclei. Each nuclear spins have two peaks in the spectrum. Expanding the resonance frequencies up to first order with respect to hyperfine coupling results that the weakly coupled nuclei keep their order as they go further away from the nuclei. This fact makes their characterization rather straight forward. To observe the weakly coupled nuclei more pronouncedly, one can use DDESEEM sequence by adding more $\pi$ pulses in the entangling periods. To distinguish other nuclei, extra measurements is required. One method is to use two dimensional hyperfine correlation spectroscopy (2D Hyscore)  \cite{vorobyov2022addressing}. However, this method is rather time-consuming as two time variables has to be swept. We suggest sweeping $\tau$ parameter and keep the track of frequency amplitude as a function of $\tau$. Two frequencies that belong to the same nuclear are correlated because the modulation depth oscillates with blind spot terms of both frequencies $\sin^2(\frac{\omega_0 \tau}{2}) \sin^2(\frac{\omega_1 \tau}{2})$. Fig. \ref{fig:5} shows two frequencies goes to bright and blind spots simultaneously. Hence, taking two dimensional correlation of the spectrum, one can deduce which two frequencies are correlated and belong to the same nuclear. \\

@@

%%%%%%%%%%%%%=========
\appendix
%%%%%%%%%%%%%=========
\section{5pESEEM formula}
\label{app:5pESEEM}
Here we need to define a few unitary operators/ The evolution in the first Hahn-echo is given by V operators, the middle free evolution is given by F operators, and the second Hahn-echo is given by W operators:
\begin{gather}
	V_0 = U_0 (\tau_1) U_1 (\tau_1),\,\,  V_1 = U_1 (\tau_1) U_0 (\tau_1) \\
	F_0 = U_0 (T), \,\, F_1 = U_1 (T) U_0 \\
	W_0 = U_0 (\tau_2) U_1 (\tau_2), \,\, W_1 = U_1 (\tau_2) U_0 (\tau_2)
\end{gather}
The signal can be obtained from four different types of trajectories that the electron spin can take. We use the fact that multiplication of unitary operators is also a unitary operator and every unitary in spin one half space can be written in terms of a rotation angle, axis, and Pauli Matrices $\vec{\sigma}$. We denote the unitary operator $U = \exp(-i \theta_U \hat{n}_U\cdot\vec{\sigma}) = M\cos(\theta_U) \mathbf{I}- i \sin(\theta_U) \hat{n}_U\cdot\vec{\sigma}$. The signal for 5-pulse ESEEM can be obtained as follows:
\begin{gather}
		\expval{\sigma_z}_{5p} =\frac{1}{4}\Big( \prod_{j=1}^n [\cos(\theta_{W_1 F_1 V_0 V_1^\dagger F_1^\dagger W_0^\dagger})]^{(j)}  \notag \\
		-\prod_{j=1}^n [\cos(\theta_{W_1 F_1 V_1 V_0^\dagger F_1^\dagger W_0^\dagger})]^{(j)}  \notag \\
		+ \prod_{j=1}^n [\cos(\theta_{W_1 F_0 V_0 V_1^\dagger F_0^\dagger W_0^\dagger})]^{(j)}  \notag \\
		-\prod_{j=1}^n [\cos(\theta_{W_1 F_0 V_1 V_0^\dagger F_0^\dagger W_0^\dagger})]^{(j)} \Big) \textcolor{gray}{e^{-\frac{2\tau_1+2\tau_2}{T_2}- \frac{T}{T_1}}}
\end{gather}
Multiplying the matrices and finding the rotation angles gives the expectation value of $\sigma_z$ for 5pESEEM sequence as it can be found in this article \cite{kasumaj20085}:
\begin{gather}
	\expval{\sigma_z}_{5p} = \frac{1}{4}\Big( \prod_{j=1}^n E_{\alpha_+}^{(j)} - \prod_{j=1}^n E_{\alpha_-}^{(j)} + \prod_{j=1}^n E_{\beta_+}^{(j)}- \prod_{j=1}^n E_{\beta_-}^{(j)}\Big)
\end{gather}
Where each term can be calculated as follows:
\small   
\begin{gather} 
	E_{\alpha_\pm}^{(k)} = E_{2p}(\tau_1) E_{2p}(\tau_2) \mp B \Big(-4k^2C_\alpha \notag \\
	+4k\cos^4(\eta) \cos(\omega_\alpha T + \phi_{\alpha_+} +\phi_{\beta_+}) 
	+2 k^2 \cos(\phi_{\beta_-}) \cos(\omega_\alpha T+\phi_{\alpha_+}) \notag \\
	+ 4k \sin^4(\eta) \cos(\omega_\alpha T +\phi_{\alpha_+} -\phi_{\beta_+} )\Big)
	\label{eq:eseem}
\end{gather}
\normalsize
$ E_{2p} $ is 2-pulse ESEEM sequence signal:
\begin{gather}
	E_{2p}(t)  = (1 - \frac{k}{2}) + \frac{k}{2} \Big(\cos(\omega_\alpha t)+\cos(\omega_\beta t) \notag \\
	-\frac{1}{2} \cos(\omega_{-}t)  -\frac{1}{2} \cos(\omega_{+}t)   \Big)
\end{gather}
With $\omega_{\pm} = \omega_\alpha \pm \omega_\beta$,  B is the blind spot term and $C_\alpha$ is a constant term:
\begin{gather}
	B = \sin(\frac{\omega_\alpha \tau_1}{2}) \sin(\frac{\omega_\alpha \tau_2}{2}) \sin(\frac{\omega_\beta \tau_1}{2}) \sin(\frac{\omega_\beta \tau_2}{2})
	\label{eq:BS}\\
	C_\alpha =\cos(\frac{\omega_\alpha \tau_1}{2}) \cos(\frac{\omega_\alpha \tau_2}{2}) \sin(\frac{\omega_\beta \tau_1}{2}) \sin(\frac{\omega_\beta \tau_2}{2}) \label{eq:Cterms}
\end{gather}
The resonance frequencies are $\omega_{\alpha(\beta)} = \sqrt{(\omega_L +s_{0(1)} A_{zz} )^2 +(s_{0(1)} A_{zx})^2}$. The quantization axis of nuclear spins is tilted by $\eta_{\alpha(\beta)} = \arctan(\frac{s_{0(1)}A_{zx}}{\omega_L+ s_{0(1)}A_{zz}})$, which gives the parameter $\eta=\frac{\eta_\alpha - \eta_\beta}{2}$. The modulation depth of each nuclear spin is $k=\sin^2(2\eta)=(\frac{(s_1-s_0)\omega_L A_{zx}}{\omega_\alpha \omega_\beta})^2$, the phase shifts are $\phi_{\alpha_\pm} = \frac{\omega_\alpha (\tau_1 \pm \tau_2)}{2}$ and $\phi_{\beta_\pm} = \frac{\omega_\beta (\tau_1 \pm \tau_2)}{2}$. The two expression for the $\beta$ pathways can be obtained by exchanging $\alpha$ and $\beta$ in equation \ref{eq:eseem}, \ref{eq:BS}, and \ref{eq:Cterms}.\\

>>>>

\title{Blueprint for nuclear spin characterization in  color center}% Force line breaks with \\
%\thanks{A footnote to the article title}%

@@

%%%%%%%%%%%%%%%%%%%================= %THEORY
\section{Methods}
\label{sec:theory}
%%%%%%%%%%%%%%%Í%%%%=================
In this section, we provide the theoretical formulas for. In a central electron spin with the electron spin $\boldmath{S}$, we only apply pulses resonant with two of the electron spin sublevels, denoted as $m_s=s_0$ and $m_s = s_1$. These two spin sublevels are separated, due to Zeeman effect or zero field splitting, and form an artificial two-level system with energy splitting of $\omega_a$. Now, we add spin $\frac{1}{2}$nuclear spins to the system (for considering P1 centers, one has to add nuclear quadrupole to the presented Hamiltonian.). The Hamiltonian of a central electron spins system surrounded by nuclear spins in the secular approximation in the rotating frame of the applied microwave $\omega_{mw}$ (considering rotating wave approximation) can be written as:
\begin{align}
	H = \Delta S_z + \sum_k (\omega_L^{(k)}  + A_{zz}^{(k)} S_z) I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
	\label{eq:H}
\end{align}
Where $\Delta = \omega_a - \omega_{mw}$ is the detuning, $\omega_L^{(k)} =\gamma_n^{(k)} B$ is nuclear Rabi frequency with $\gamma_n^{(k)}$ nuclear spin gyromagnetic ratio and $B$ external magnetic field, $A_{zz}, A_{zx}$ parallel and perpendicular secular component of hyperfine tensor. This Hamiltonian results that depending on the electron spin sublevel $s_i$ ($i=0,1$), the nuclear spin rotate with the resonance frequency of $\omega_i = \sqrt{(\omega_L + s_i A_{zz})^2+ (s_i A_{zx})^2}$, $i=0,1$ along the axes $\vec{n}_i =(\frac{s_i A_{zx}}{\omega_i}, 0, \frac{\omega_L + s_i A_{zz}} {\omega_i})$. It is assumed that the pulse duration $t_p$ are short enough ($\omega_L t_p \ll 1$) that the nuclear spin dynamics during this time is negligible. The most simplest characterization sequence is Ramsey. \\ The Ramsey signal can be obtained as follows:
\begin{gather}
	\expval{\sigma_z}_{Ram} = \cos(\Delta \tau) \prod_{k=1}^{n} \Big[ \cos(\frac{\omega_0 \tau}{2}) \cos(\frac{\omega_1 \tau}{2})+ \notag \\
	(\vec{n_0}\cdot \vec{n_1})\sin(\frac{\omega_0 \tau}{2}) \sin(\frac{\omega_1 \tau}{2}) \Big]^{(k)}
	\textcolor{gray}{e^{-\frac{\tau}{T_2^\ast}}}
\end{gather} 
We do not take any relaxation process into account and the gray exponential decay is put by hand to remember the relaxation time scale. Ramsey signal decay with electron spin $T_2^\ast$, so it can be used to observe the core of the defect. Ramsey sequence is sensitive to detuning since the electron spin is not refocused, however it is possible to sense nuclear spins with vanishing perpendicular hyperfine coupling.\\

@@

		\includegraphics[width=0.8\textwidth]{pict/dd_compare2.pdf}

@@

We want to make a sequence that is limited by T1 instead of T2. The idea of 5-pulse ESEEM \cite{kasumaj20085} is that to create an entanglement (Hahn-echo) before the free evolution (Ramsey). In other words, change the initial density matrix for the Ramsey sequence from such that entanglement already exist in the electron polarization terms of the density matrix. Fig \ref{fig:seqs} shows the sequence, which can be interpreted as two Hahn Echos that are separated by a long free evolution $T\gg T_2^\ast$ such that the coherence of the electron spin vanishes. The analytical formula to this sequence is provided in the appendix \ref{app:5pESEEM}. During the free evolution time, each nuclear spins oscillates with either of two resonance frequencies obtained from eq. \ref{eq:H}.
\begin{figure*}[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{pict/eseem_compare2.pdf}
		\caption{a. FFT of ESEEM signal b. a zoom into weakly coupled nuclear spins c. Enhancing weakly coupled nuclear spins by having 72 pulses in each entangling period. $\tau$ is set to the Larmor bright spot.}
		\label{fig:4}
	\end{center}
\end{figure*}
Fig \ref{fig:4} shows the furrier transform of the signal. Zooming into area close to the Larmor frequency reveals the weakly coupled nuclei. Each nuclear spins have two peaks in the spectrum. Expanding the resonance frequencies up to first order with respect to hyperfine coupling results that the weakly coupled nuclei keep their order as they go further away from the nuclei. This fact makes their characterization rather straight forward. To observe the weakly coupled nuclei more pronouncedly, one can use DDESEEM sequence by adding more $\pi$ pulses in the entangling periods. To distinguish other nuclei, extra measurements is required. One method is to use two dimensional hyperfine correlation spectroscopy (2D Hyscore)  \cite{vorobyov2022addressing}. However, this method is rather time-consuming as two time variables has to be swept. We suggest sweeping $\tau$ parameter and keep the track of frequency amplitude as a function of $\tau$. Two frequencies that belong to the same nuclear are correlated because the modulation depth oscillates with blind spot terms of both frequencies $\sin^2(\frac{\omega_0 \tau}{2}) \sin^2(\frac{\omega_1 \tau}{2})$. Fig. \ref{fig:5} shows two frequencies goes to bright and blind spots simultaneously. Hence, taking two dimensional correlation of the spectrum, one can deduce which two frequencies are correlated and belong to the same nuclear. \\

@@

%%%%%%%%%%%%%=========
\appendix
%%%%%%%%%%%%%=========
\section{5pESEEM formula}

Here are the sequences:
\label{app:5pESEEM}
Here we need to define a few unitary operators/ The evolution in the first Hahn-echo is given by V operators, the middle free evolution is given by F operators, and the second Hahn-echo is given by W operators:
	V_0 = U_0 (\tau_1) U_1 (\tau_1),\,\,  V_1 = U_1 (\tau_1) U_0 (\tau_1) \\
	F_0 = U_0 (T), \,\, F_1 = U_1 (T) U_0 \\
	W_0 = U_0 (\tau_2) U_1 (\tau_2), \,\, W_1 = U_1 (\tau_2) U_0 (\tau_2)
The signal can be obtained from four different types of trajectories that the electron spin can take. We use the fact that multiplication of unitary operators is also a unitary operator and every unitary in spin one half space can be written in terms of a rotation angle, axis, and Pauli Matrices $\vec{\sigma}$. We denote the unitary operator $U = \exp(-i \theta_U \hat{n}_U\cdot\vec{\sigma}) = M\cos(\theta_U) \mathbf{I}- i \sin(\theta_U) \hat{n}_U\cdot\vec{\sigma}$. The signal for 5-pulse ESEEM can be obtained as follows:
		\expval{\sigma_z}_{5p} =\frac{1}{4}\Big( \prod_{j=1}^n [\cos(\theta_{W_1 F_1 V_0 V_1^\dagger F_1^\dagger W_0^\dagger})]^{(j)}  \notag \\
		-\prod_{j=1}^n [\cos(\theta_{W_1 F_1 V_1 V_0^\dagger F_1^\dagger W_0^\dagger})]^{(j)}  \notag \\
		+ \prod_{j=1}^n [\cos(\theta_{W_1 F_0 V_0 V_1^\dagger F_0^\dagger W_0^\dagger})]^{(j)}  \notag \\
		-\prod_{j=1}^n [\cos(\theta_{W_1 F_0 V_1 V_0^\dagger F_0^\dagger W_0^\dagger})]^{(j)} \Big) \textcolor{gray}{e^{-\frac{2\tau_1+2\tau_2}{T_2}- \frac{T}{T_1}}}
Multiplying the matrices and finding the rotation angles gives the expectation value of $\sigma_z$ for 5pESEEM sequence as it can be found in this article \cite{kasumaj20085}:
	\expval{\sigma_z}_{5p} = \frac{1}{4}\Big( \prod_{j=1}^n E_{\alpha_+}^{(j)} - \prod_{j=1}^n E_{\alpha_-}^{(j)} + \prod_{j=1}^n E_{\beta_+}^{(j)}- \prod_{j=1}^n E_{\beta_-}^{(j)}\Big)
Where each term can be calculated as follows:
\small   
	E_{\alpha_\pm}^{(k)} = E_{2p}(\tau_1) E_{2p}(\tau_2) \mp B \Big(-4k^2C_\alpha \notag \\
	+4k\cos^4(\eta) \cos(\omega_\alpha T + \phi_{\alpha_+} +\phi_{\beta_+}) 
	+2 k^2 \cos(\phi_{\beta_-}) \cos(\omega_\alpha T+\phi_{\alpha_+}) \notag \\
	+ 4k \sin^4(\eta) \cos(\omega_\alpha T +\phi_{\alpha_+} -\phi_{\beta_+} )\Big)
	\label{eq:eseem}
\normalsize
$ E_{2p} $ is 2-pulse ESEEM sequence signal:
	E_{2p}(t)  = (1 - \frac{k}{2}) + \frac{k}{2} \Big(\cos(\omega_\alpha t)+\cos(\omega_\beta t) \notag \\
	-\frac{1}{2} \cos(\omega_{-}t)  -\frac{1}{2} \cos(\omega_{+}t)   \Big)
With $\omega_{\pm} = \omega_\alpha \pm \omega_\beta$,  B is the blind spot term and $C_\alpha$ is a constant term:
	B = \sin(\frac{\omega_\alpha \tau_1}{2}) \sin(\frac{\omega_\alpha \tau_2}{2}) \sin(\frac{\omega_\beta \tau_1}{2}) \sin(\frac{\omega_\beta \tau_2}{2})
	\label{eq:BS}\\
	C_\alpha =\cos(\frac{\omega_\alpha \tau_1}{2}) \cos(\frac{\omega_\alpha \tau_2}{2}) \sin(\frac{\omega_\beta \tau_1}{2}) \sin(\frac{\omega_\beta \tau_2}{2}) \label{eq:Cterms}
The resonance frequencies are $\omega_{\alpha(\beta)} = \sqrt{(\omega_L +s_{0(1)} A_{zz} )^2 +(s_{0(1)} A_{zx})^2}$. The quantization axis of nuclear spins is tilted by $\eta_{\alpha(\beta)} = \arctan(\frac{s_{0(1)}A_{zx}}{\omega_L+ s_{0(1)}A_{zz}})$, which gives the parameter $\eta=\frac{\eta_\alpha - \eta_\beta}{2}$. The modulation depth of each nuclear spin is $k=\sin^2(2\eta)=(\frac{(s_1-s_0)\omega_L A_{zx}}{\omega_\alpha \omega_\beta})^2$, the phase shifts are $\phi_{\alpha_\pm} = \frac{\omega_\alpha (\tau_1 \pm \tau_2)}{2}$ and $\phi_{\beta_\pm} = \frac{\omega_\beta (\tau_1 \pm \tau_2)}{2}$. The two expression for the $\beta$ pathways can be obtained by exchanging $\alpha$ and $\beta$ in equation \ref{eq:eseem}, \ref{eq:BS}, and \ref{eq:Cterms}.\\
